Supplementary material

### 2020-04-23 analysis/paper/paper.Rmd
- More changes to paper and supplementary

Data collection was done through E4, a software developed by Dibble and McPherron (2003). The complete Data Dictionary with all variables, attributes, values and description, with reference to the consulted literature is available in our research compendium at https://doi.org/xxxx, and can be consulted in the supplementary material (Supplementary table 1).

@@

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (Portugal), near a small homonymous village, within the municipality of Vila do Bispo (in the Supplementary Material). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, with a natural boundary to the east and bordered by a limestone hill through all its extension. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest [@bicho2003; @cascalheira2010; @cascalheiraetal2008].

@@

Since 2012, six layers have been identified (Supplementary table 2). In some of these layers, lateral sediment variations were identified, which were coded in the field through the concatenation of a letter to the layer number (*e.g.* 4E). In many of these cases, the isolation of this vertical variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

@@

These layers, 4E and 5, are vertically contiguous, separated by a relatively flat surface covered by large limestone blocks. Unlike what was expected, however, these levels did not reveal a Gravettian assemblage with characteristics similar to previous years [@marreiros2009], but rather materials with patterns similar to those expected in a Proto-Solutrean assemblage (the most noticeable of all being the Vale Comprido points and blanks). The context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (complete table with calibrated radiocarbon dates may be found in the Supplementary Materials).

@@

One of the most noticeable features about the vertical distribution of pieces in the Terrace (row H and I; Supplementary Figure 4) is the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with the higher concentration of lithic materials in these layers. Likewise, the presence of Vale Comprido points is also associated with Layer 4E and the top levels of Layer 5 [@belmiro2017], which seems to occur around 25 ka cal BP (Figure \@ref(fig:c14VB)). The WK-44416 date seems to show that, at least from around c. 26 ka cal BP, the site was occupied even if with lower intensity, a date that falls onto the Final Gravettian/Proto-Solutrean or Final Gravettian/Terminal Gravettian transition in the traditional models [@zilhao1997].

@@

A total of 26,703 pieces were analysed for both groups, 11,094 from Lower 5 group and 15,609 for the Upper 5/4E (both tables may be found in the Supplementary Materials). Most of these are chips (70.62% from Lower 5 and 65.68% from Upper 5/4E), followed by shatters, which make up c. 21% and c. 24% of the groups, respectively. These extremely high numbers for debris are mostly the result of quartz use, which can be explained not only by on-site knapping of this raw material but mostly by its breakage patterns (especially when coarser), which typically produces more waste than in other raw materials.

@@

During the analysis, an unprecedented type of raw material in the whole site of Vale Boi was identified.  Dolerite seems to appear only in layers 4E and 5 of the Terrace and Proto-Solutrean levels of the Slope area. The most striking characteristic of this raw material in the assemblages under study is the low presence of chips or shatter (n=1 each), but high frequencies of blanks (50% in Lower 5 and 60.8% in Upper 5/4E) and blank fragments (16.6% in Lower 5 and 13% in Upper 5/4E). Unlike Lower 5, in Upper 5/4E dolerite is represented by five retouched tools, some of which are Vale Comprido points. This raw material seems to have been used preferentially at Vale Boi for the production of these types of points, a pattern already observed in previous works [@marreiros2009], where three out of five Vale Comprido points were identified as dolerite (originally jasper).

@@

Excluding inform cores,there is a clear dominance of single platform cores, for most raw materials and in both phases. Chert presents the highest variability of core types, with high frequencies for single platform (50% for Lower 5 and 33.3% in Upper 5/4E), but also prismatic (37.5% on Lower 5). On Upper 5/4E, there are also opposed, pyramidal, two single platforms and other types of platforms in percentages higher than 10%. In the Lower 5 group chert unidirectional prismatic cores are more frequent.

Most of the cores were used for the extraction of flakes, although blade and bladelet scars also show relatively high frequencies associated with the exploitation of unidirectional prismatic cores in Lower 5 and Upper 5/4E groups.

Most of the analysed core platforms are plain or cortical. On Upper 5/4E there is a small frequency of faceted platforms on both quartz and chert (3.4% and 18.5% respectively). Platforms width and thickness means show smaller platforms for chert on both phases, while greywacke shows the highest means for platform measurements. This pattern is similar for other measurements, for which chert and quartz exhibit the smaller means.

Regarding flakes, morphologically, Lower 5 flakes seem to have mostly irregular shapes, with frequencies over 45% for most raw materials, followed by convergent and parallel shapes in much lower frequencies (under 15%). Dolerite, unlike other raw materials, shows only convergent and parallel shapes even if with low representativity (n=3). Cross sections are mostly irregular and triangular, varying slightly in frequency regarding each raw material. All raw materials, with exception of quartz, show high frequencies of straight profiles (over 40%), chert showing a similar percentage between straight and curved profiles (44.3% and 42.9% respectively). Quartz, however, has a frequency of 65% of straight profiles, showing much smaller frequencies for other types.

@@

Elongated blanks have mostly parallel and convergent shapes, on both phases, although Lower 5 shows significant differences across the different raw materials, since c. 64% of quartz has parallel shapes, whereas c. 52% of chert has convergent ones. Cross sections are mostly triangular, with straight and curved profiles. Differences in profile types are rather significant between both phases, but only for quartz artifacts. While in Lower 5, quartz elongated blanks present a percentage of curved profiles of c. 23%, in Upper 5/4E, this percentage drops to c. 12%. Curved and twisted profiles in bladelets are usually associated with the exploitation of carinated cores, particularly in the Gravettian-Solutrean transition as presented by @almeida2000 for the Terminal Gravettian of Estremadura. Although no carinated cores were identified in the Lower 5 assemblage, the presence of a significant number of bladelets with curved profiles might indicate the use of such a type of reduction strategy.

@@

As seen before, retouched pieces make up a small percentage of the assemblage, with 167 identified pieces, from which 55 can be inserted in the Lower 5 group, while 112 in the Upper 5/4E group.

@@

The interior of the chamber is about 11 x 14 m, and has more than 10 m of coarse sedimentary infill in inclined beds, derived from roof collapse, gravity flows and fine sediment infiltration (Supplementary Figure 9) [@benedettietal2019]. The cave opening is marked by the existence of a large limestone cone of stone blocks [@bicho2006] while the sediment inside the cave consists of smaller and angular limestone clasts in a matrix of fine sediment [@benedettietal2019].

@@

A complete description of all currently identified levels in the site and, whenever existent, associated lithic assemblages, may be found in the Supplementary Materials. From the 34 levels described in the table, 23 show human occupations or association with a lithic assemblage, and 20 of these can be attributed to the Upper Paleolithic:

@@

For U and lower T, radiocarbon dates (Supplementary Table 3), all obtained from charcoal samples, show results of c. 27-26 cal ka BP, except for one of lower T dates which presents a result of c. 25-24 cal ka BP.

@@

There is a clear separation between the two groups, one composed of levels U and lower T, and the other of middle T. This is also visible on the spatial dispersion of lithics and other artifacts (Supplementary Figure 11), with an accumulation on the bottom left, correspondent to the U/T levels, and another aggregation in the middle portion, to the right, ranging from 20-40 cm of depth difference between the groups.

@@

A total of 376 pieces were analysed, 196 coming from the U/Lower T phase and 180 from the Middle T group. In both groups, debitage waste is mostly composed of chips, which represents 49.5% of the U/Lower T group and 37.2% of the Middle T group. As in Vale Boi, these values for chippage can be mostly explained by quartz breakage patterns.

@@

A total of 7 cores were recorded on the Lapa do Picareiro assemblages, two in the U/Lower T group (both in quartz) and five in the Middle T, of which 2 are in chert and 2 are in quartz.

@@

There are a total of plotted 122 flakes, 46 attributed to the U/Lower T group, more than 50% being in quartz, and 67 attributed to the Middle T group, its vast majority in chert. On both groups, attributes are very similar.

@@

In fact, both blades and bladelets show a majority of parallel and convergent edge shapes, straight and curved profiles. Dorsal scars are unidirectional on the dorsal surfaces of all artifacts, with a majority of two previous removals for quartz and two to four scars for chert. Platforms are mostly plain, with no cortex.

The total number of retouched pieces from Lapa do Picareiro with spatial information, allowing the identification of phase is 11, of which 6 belong to the U/Lower T group and 5 to the Middle T group.

>>>>

Data collection was done through E4, a software developed by Dibble and McPherron (2003). The complete Data Dictionary with all variables, attributes, values and description, with reference to the consulted literature is available in our research compendium at https://doi.org/xxxx, and can be consulted in the supplementary material (Table S1).

@@

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (Portugal), near a small homonymous village, within the municipality of Vila do Bispo (Figure S2 and S3). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, with a natural boundary to the east and bordered by a limestone hill through all its extension. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest [@bicho2003; @cascalheira2010; @cascalheiraetal2008].

@@

Since 2012, six layers have been identified (Table S2 in the Supplementary Materials). In some of these layers, lateral sediment variations were identified, which were coded in the field through the concatenation of a letter to the layer number (*e.g.* 4E). In many of these cases, the isolation of this vertical variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

@@

These layers, 4E and 5, are vertically contiguous, separated by a relatively flat surface covered by large limestone blocks. Unlike what was expected, however, these levels did not reveal a Gravettian assemblage with characteristics similar to previous years [@marreiros2009], but rather materials with patterns similar to those expected in a Proto-Solutrean assemblage (the most noticeable of all being the Vale Comprido points and blanks). The context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (Table S3).

@@

One of the most noticeable features about the vertical distribution of pieces in the Terrace (row H and I; SFigure S4) is the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with the higher concentration of lithic materials in these layers. Likewise, the presence of Vale Comprido points is also associated with Layer 4E and the top levels of Layer 5 [@belmiro2017], which seems to occur around 25 ka cal BP (Figure \@ref(fig:c14VB)). The WK-44416 date seems to show that, at least from around c. 26 ka cal BP, the site was occupied even if with lower intensity, a date that falls onto the Final Gravettian/Proto-Solutrean or Final Gravettian/Terminal Gravettian transition in the traditional models [@zilhao1997].

@@

A total of 26,703 pieces were analysed for both groups, 11,094 from Lower 5 group and 15,609 for the Upper 5/4E (Tables S4 and S5). Most of these are chips (70.62% from Lower 5 and 65.68% from Upper 5/4E), followed by shatters, which make up c. 21% and c. 24% of the groups, respectively. These extremely high numbers for debris are mostly the result of quartz use, which can be explained not only by on-site knapping of this raw material but mostly by its breakage patterns (especially when coarser), which typically produces more waste than in other raw materials.

@@

During the analysis, an unprecedented type of raw material in the whole site of Vale Boi was identified (Figure S5).  Dolerite seems to appear only in layers 4E and 5 of the Terrace and Proto-Solutrean levels of the Slope area. The most striking characteristic of this raw material in the assemblages under study is the low presence of chips or shatter (n=1 each), but high frequencies of blanks (50% in Lower 5 and 60.8% in Upper 5/4E) and blank fragments (16.6% in Lower 5 and 13% in Upper 5/4E). Unlike Lower 5, in Upper 5/4E dolerite is represented by five retouched tools, some of which are Vale Comprido points. This raw material seems to have been used preferentially at Vale Boi for the production of these types of points, a pattern already observed in previous works [@marreiros2009], where three out of five Vale Comprido points were identified as dolerite (originally jasper).

@@

Excluding inform cores,there is a clear dominance of single platform cores, for most raw materials and in both phases. Chert presents the highest variability of core types, with high frequencies for single platform (50% for Lower 5 and 33.3% in Upper 5/4E), but also prismatic (37.5% on Lower 5). On Upper 5/4E, there are also opposed, pyramidal, two single platforms and other types of platforms in percentages higher than 10%. In the Lower 5 group chert unidirectional prismatic cores are more frequent (Tables S6 and S7).

Most of the cores were used for the extraction of flakes, although blade and bladelet scars also show relatively high frequencies associated with the exploitation of unidirectional prismatic cores in Lower 5 and Upper 5/4E groups (Figure S6).

Most of the analysed core platforms are plain or cortical. On Upper 5/4E there is a small frequency of faceted platforms on both quartz and chert (3.4% and 18.5% respectively). Platforms width and thickness means show smaller platforms for chert on both phases, while greywacke shows the highest means for platform measurements. This pattern is similar for other measurements, for which chert and quartz exhibit the smaller means (Figure S7, Tables S8 and S9).

Regarding flakes (Tables S10 and S11), morphologically, Lower 5 flakes seem to have mostly irregular shapes, with frequencies over 45% for most raw materials, followed by convergent and parallel shapes in much lower frequencies (under 15%). Dolerite, unlike other raw materials, shows only convergent and parallel shapes even if with low representativity (n=3). Cross sections are mostly irregular and triangular, varying slightly in frequency regarding each raw material. All raw materials, with exception of quartz, show high frequencies of straight profiles (over 40%), chert showing a similar percentage between straight and curved profiles (44.3% and 42.9% respectively). Quartz, however, has a frequency of 65% of straight profiles, showing much smaller frequencies for other types.

@@

Elongated blanks have mostly parallel and convergent shapes, on both phases, although Lower 5 shows significant differences across the different raw materials, since c. 64% of quartz has parallel shapes, whereas c. 52% of chert has convergent ones. Cross sections are mostly triangular, with straight and curved profiles (Tables S14 and S15). Differences in profile types are rather significant between both phases, but only for quartz artifacts. While in Lower 5, quartz elongated blanks present a percentage of curved profiles of c. 23%, in Upper 5/4E, this percentage drops to c. 12%. Curved and twisted profiles in bladelets are usually associated with the exploitation of carinated cores, particularly in the Gravettian-Solutrean transition as presented by @almeida2000 for the Terminal Gravettian of Estremadura. Although no carinated cores were identified in the Lower 5 assemblage, the presence of a significant number of bladelets with curved profiles might indicate the use of such a type of reduction strategy.

@@

As seen before, retouched pieces make up a small percentage of the assemblage, with 167 identified pieces, from which 55 can be inserted in the Lower 5 group, while 112 in the Upper 5/4E group (Tables S18 and S19.

@@

The interior of the chamber is about 11 x 14 m, and has more than 10 m of coarse sedimentary infill in inclined beds, derived from roof collapse, gravity flows and fine sediment infiltration (Figure S9 and S10) [@benedettietal2019]. The cave opening is marked by the existence of a large limestone cone of stone blocks [@bicho2006] while the sediment inside the cave consists of smaller and angular limestone clasts in a matrix of fine sediment [@benedettietal2019].

@@

A complete description of all currently identified levels in the site and, whenever existent, associated lithic assemblages, may be found in the Supplementary Materials (Table S20). From the 34 levels described in the table, 23 show human occupations or association with a lithic assemblage, and 20 of these can be attributed to the Upper Paleolithic:

@@

For U and lower T, radiocarbon dates (Table S3), all obtained from charcoal samples, show results of c. 27-26 cal ka BP, except for one of lower T dates which presents a result of c. 25-24 cal ka BP.

@@

There is a clear separation between the two groups, one composed of levels U and lower T, and the other of middle T. This is also visible on the spatial dispersion of lithics and other artifacts (Figure S11), with an accumulation on the bottom left, correspondent to the U/T levels, and another aggregation in the middle portion, to the right, ranging from 20-40 cm of depth difference between the groups.

@@

A total of 376 pieces were analysed, 196 coming from the U/Lower T phase and 180 from the Middle T group (Tables S21 and S22). In both groups, debitage waste is mostly composed of chips, which represents 49.5% of the U/Lower T group and 37.2% of the Middle T group. As in Vale Boi, these values for chippage can be mostly explained by quartz breakage patterns.

@@

A total of 7 cores were recorded on the Lapa do Picareiro assemblages, two in the U/Lower T group (both in quartz) and five in the Middle T, of which 2 are in chert and 2 are in quartz (Table S23).

@@

There are a total of plotted 122 flakes, 46 attributed to the U/Lower T group, more than 50% being in quartz, and 67 attributed to the Middle T group, its vast majority in chert. On both groups, attributes are very similar (Tables S27 and S28).

@@

In fact, both blades and bladelets show a majority of parallel and convergent edge shapes, straight and curved profiles. Dorsal scars are unidirectional on the dorsal surfaces of all artifacts, with a majority of two previous removals for quartz and two to four scars for chert. Platforms are mostly plain, with no cortex (Tables S31 and S32).

The total number of retouched pieces from Lapa do Picareiro with spatial information, allowing the identification of phase is 11, of which 6 belong to the U/Lower T group and 5 to the Middle T group (Tables S35 and S36).

### 2020-04-23 analysis/paper/supplementary.Rmd
- More changes to paper and supplementary

---
title: "Supplementary material"
header-includes:
  - \usepackage{caption}
  - \captionsetup[figure]{name=Supplementary Figure,labelfont=it,labelsep = period}
  - \captionsetup[table]{name=Supplementary Table,labelfont=it,labelsep = period}  
output: 
  pdf_document: 
    latex_engine: xelatex

---

>>>>

---
title: "Supplementary material"
output: 
  pdf_document: 
    latex_engine: xelatex
header-includes: 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

---
\beginsupplement
